---
title: "Analysis-Checklist"
author: "Paige Magoto"
format: html
editor: visual
---

## Exploratory Data Analysis Checklist:

1.  Formulate your question

    a\. Which state was the most common birthplace for judges in this data set?

    b\. Whats the most common reason for termination, if any?

    c\. Number of male vs female?

2.  Read in your data

    ```{r}
    library(DigitalMethodsData)
    data("judges")
    library(tidyverse)
    ```

3.  Check the packaging

    ```{r}
    nrow(judges)
    ncol(judges)
    ```

4.  Run `str()`

    ```{r}
    str(judges)
    ```

5.  Look at the top and the bottom of your data

    ```{r}
    head(judges)
    tail(judges)
    ```

6.  Check your “n”s

    ```{r}
    judge.state <- judges %>% select(Last.Name, First.Name, Middle.Name, Birth.State) 
    select(judges, Birth.State) %>% nrow
    unique(judges$Birth.State)
    ```

    ```{r}
    judge.termination <- judges %>% select(Last.Name, First.Name, Middle.Name, Termination..1., Termination..2., Termination..3., Termination..4., Termination..5., Termination..6.)
    unique(judges$Termination..1.)
    unique(judges$Termination..2.)
    unique(judges$Termination..3.)
    unique(judges$Termination..4.)
    unique(judges$Termination..5.)
    unique(judges$Termination..6.)
    select(judges,Termination..1., Termination..2., Termination..3., Termination..4., Termination..5., Termination..6.)

    ```

    ```{r}
    judge.gender <- judges %>% select(Last.Name, First.Name, Middle.Name, Gender)
    unique(judges$Gender)
    ```

7.  Validate with at least one external data source

    I goggled a few of the Judges birth places.

8.  Try the easy solution first

    ```{r}
    #state/region
    state.rank<- judge.state %>% group_by(Birth.State) %>% summarise(count =n()) %>% arrange(count)
    # reason for termination 
    term.rank <- judge.termination %>% group_by(Termination..1., Termination..2., Termination..3., Termination..4., Termination..5., Termination..6.) %>% summarise(count = n())
    #does not work above^^
    #gender
    gender.judge <- judge.gender %>% group_by(Gender) %>% summarise(count = n())
    ```

9.  Challenge your solution

    ```{r}
    gender.counts <- judges %>% group_by(Birth.Year, Gender) %>% summarise(count = n()) %>% pivot_wider(names_from = Gender, values_from = count, values_fill = 0)
    #This data frame uses similar ideas from before, but gives mroe valueable information. 
    ```
